Recruiting Chapter

### 2020-10-13 manuscript.Rmd
working on manuscript draft

We present this chapter as one example of empirical research about enrollment management behavior. When we contrast research about enrollment management to descriptions of enrollment management practice, from investigative journalists or from consulting firms [cite], it becomes apparent that the research community is decades behind practice and has ignored a great number of enrollment management practices that affect access to higher education [AN EXCEPTION TO THE CRITIQUE IS A SMALL BUT HIGH-QUALITY CASE STUDY LITERATURE FROM SOCIOLOGY]. Thus, the enrollment management remains an opaque industry to researchers and, in turn, to policymakers and the public. Developing thoughtful state and federal policies about enrollment management depends on researchers doing the slow, methodical work of developing an empirical literature that documents enrollment management practices and evaluates their effect on opportunities for students. 
[TONE DOWN THE CRITIQUE; DON'T WANT TO MAKE IT SEEM LIKE WE THINK WE ARE THE FIRST TO STUDY THIS STUFF; GIVE CREDIT TO THE EXCELLENT CASE STUDY RESEARCH]

The "enrollment funnel," a conceptual heuristic used by the enrollment management industry and shown in figure X, is helpful for framing the development of this empirical literature. The enrollment funnel describes the stages in the student recruitment process (e.g., prospects, inquiries, applicants, accepted applicants, enrolled) in order to inform targeted recruiting interventions. The majority of research on enrollment management focuses on the final stages of the enrollment funnel, specifically which applicants are admitted and the use of financial aid "leveraging" to convert admits to enrollees.<!-- [@RN3536; @RN3544; @RN3523; @RN4131; @RN2241; @RN3564; @RN1948] -->By contrast, the enrollment management industry expends substantial resources on earlier stages of the funnel [@RN4664]. “Prospects” are “all the potential students you would want to attract to your institution” [@RN4322]. “Inquiries” are prospects that contact the institution, including those who respond to a solicitation and those who reach out on their own.

[REVISE BELOW PARAGRAPH TO TAKE OUT MOST OF THE NUMBERS OF PERCENT SPENT ON EACH ACTIVITY; BOGS DOWN THE PROSE; MAIN POINT IS HIGHLIGHTING THE ACTIVITIES THAT TARGET SPECIFIC ELEMENTS OF THE FUNNEL; READER CAN READ THE PERCENTAGES ON THEIR OWN]
Table X shows the percentage of undergraduate recruitment budget allocated to different marketing and recruiting activities. ^[table drawn directly from @RN4664 figure 9; based on a convenience sample of 45 four-year non-profit institutions and 21 four-year public institutions]. Institutions identify undergradaute prospects largely by purchasing "student lists" from College Board, ACT, and other vendors.  Table X indicates that private and public institutions spent 14% and 12%, respectively, of their undergraduate recruitment budget on student list purchases. Institutions utilize digital advertising, traditional advertising, and social media to solicit inquiries and to create positive "buzz" amongst potential stealth applicants [@RN4411][CITE OTHER]. <!-- Table X indicates that...  [summarize expenditure on stuff related to soliciting inquiries once you figure out answers to below questions] -->  Once identified, prospects/inquiries are targeted with face-to-face and remote recruiting interventions designed to solit applications and deepen engagement. Remote recruiting interventions include postcards, brochures, email, and text messages. <!-- Table X indicates that these "prospective student communications" account for 13% of the recruiting budget at private institutions and 17% of the budget at public institutions.  -->Face-to-face recruiting interventions include on-campus visits (12% and 11% of budget, respectively, at private and public institutions) and off-campus recruiting events, often referred to as "travel" (17% and 16% of budget, respectively, at private and public institutions).
<!-- QUESTIONS ABOUT THE RNL 2020 COST OF RECRUITING AN UNDERGRADUATE STUDENT REPORT, FIGURE 9

What does "recruitment publications" include?
what does "prospective student communications" include?
what does "event" include? can I assume this is on-campus events only cuz "travel" is a separate category?
what does "web services and digital advertising" include? Does this include social media?
-->
Having identified prominent marketing and recruiting activities (e.g., student list purchases, digital advertising, off-campus recruiting visits), why have so few researchers studied higher education marketing and recruitng? One challenge is that empirical research depends on data but obtaining concrete data about marketing and recruiting behaviors is difficult, especially for conventional data collection methods like surveys and interviews. Thus, empirical research requires different approaches to data collection, including public records requests and "data science" methodologies, such as web-scraping and streaming social media data.

This chapter analyzes off-campus recruiting visits made in 2017 by a sample of public research universities and a sample of selective private colleges and universities.  We collected visit data by "scraping" URLs (e.g., "Coming to a neighborhood near you") from admissions websites. Pages were scraped once per week for the entire 2017 calendar year. In addition, for the sample of public research universities, we also issued public records requests.

To date, our analyses have focused on recruiting visits by public research universities. As described below, the majority of public universities in our sample made far more out-of-state visits than in-state visits. These out-of-state visits focused on wealthy, predominantly white communities, with a disproportionate number of visits to private high schools. This chapter analyzes visits to private high schools. Using social network methods, we analyze visits by private colleges and universities to private high schools, visits by public research universities to private high schools, and then compare similarities and differences between public and private postsecondary institutions [as a means of saying something about "privatization"?]
[MOTIVATE PARTICULAR FOCUS OF THIS CHAPTER: NETWORK ANALYSIS OF VISITS TO PRIVATE HIGH SCHOOLS BY BOTH PUBLIC AND PRIVATE COLLEGES/UNIVERSITIES]

@@

@@

## Why Network Analysis and Network Basics

We utilize network methods rather than conventional descriptive statistics for two, related reasons. First, the theory and analytic methods of social network analysis privileges the relationship between actors. The discussion of Stevens and Khan suggests that for private school students, "it is not just the quality of the students that gets them into college but the quality of the relationship between elite high schools and colleges" [@RN4407] [pg. 175]. Second, a visit from a college/university admissions counselor to a high school can be conceived as a "network tie" connecting two actors in a social network.

We briefly define essential social network analysis concepts and their operationalization in our analyses.  A social network consists of a set of actors -- referred to as "vertices" -- and the connections -- referred to as "edges" between these actors.  Whereas "one mode" networks consist of vertices of the same "type" (e.g., in a publication network each vertex is an author), "two mode" networks consit of vertices associated with one type of actor/entity having connections to vertices of another type. For example, an an actor-movie network consists of actors -- mode 1 -- who appear in movies -- mode 2 and an actor shares an edge with a movie if the actor appears in the movie.  The social network analyzed in this chapter is a two-mode network, where vertices consists of colleges/universities (mode 1) and high schools (mode 2) and an "edge" is defined as a visit from a college/university to a private high school.

Our analyses draw from analyses of "corporate board-director" networks because these are the most commonly analyzed two-mode networks in the social networks literature. Board-director networks consist of directors (mode 1) who serve on organizational boards (mode 2). @RN4663 states that "Overlapping groups such as boards of directors form a two-mode membership network in which one can conceive of directors as nodes [vertices] connected by a tie of common board membership, or boards as nodes connected by a tie of one or more shared directors" [pg. xx].  Similarly, our college-school network can be conceive of colleges as vertices that are connected to high school verticies by a tie of an off-campus recruiting visit.  In practice, two-mode networks are ofen analyzed as one-mode networks. In the corporate board interlock literature a one-mode director-by-director matrix is created by identifying whether two directors serve on one ore more board (e.g., two directors share a tie if both serve on the board of Google) and a one-mode company-by-company matrix is created by identifying whether two boards have one or more directors who serve on both boards (e.g., xxxxxx). Similarly, we create a one-mode college network in which two colleges share a connection if both colleges visit at least one high school in common and a one-mode high school network in which two high schools share an edge if they are visited by at least one college/university in common.

A two-mode board-director network can also be conceived as two one-mode networks.

In the social network analysis literature, the corporate board-director network is most commonly analyzed two-mode network [CITE]. These networks consist of directors (mode 1) who serve on organizational boards (mode 2).

The most commonly analyzed two-mode network is the "corporate board interlock."

For example, in a two-mode actor-movie network consists of two modes of vertices, actors and movies.

>>>>

<!-- The "enrollment funnel," shown in Figure X, describes stages in the student recruitment process (e.g., prospects, inquiries, applicants, accepted applicants, and enrolled students). -->

Figure X shows the "enrollment funnel," a conceptual heurisitc used by the enrollment management industry, which describe the stages in the student recruitment process (e.g., prospects, inquiries, applicants, accepted applicants, and enrolled students). For example, “prospects” are “all the potential students you would want to attract to your institution” [@RN4322]. “Inquiries” are prospects that contact the institution, including those who respond to a solicitation and those who reach out on their own. The enrollment management industry uses the enrollment funnel to inform interventions that target specific stages of the funnel.

Table X, reproduced from @RN4664, shows the the percentage of undergraduate recruitment budget allocated to different marketing and recruiting activities. ^[table drawn directly from @RN4664 figure 9; based on a convenience sample of 45 four-year non-profit institutions and 21 four-year public institutions]. Institutions identify undergradaute prospects largely by purchasing "student lists" from College Board, ACT, and other vendors. Institutions utilize advertising (traditional and digital) and social media to solicit inquiries and to create positive "buzz" amongst prospects [@RN4411][CITE OTHER].  Given the the rise in “stealth applicants” who do not inquire before applying (Dupaul & Harris, 2012), advertising and social media enables universities to tell their story to prospects who do not wish to be contacted. Once prospects and inquiries are identified, they are targeted with face-to-face and remote recruiting interventions designed to solit applications and deepen engagement. Face-to-face recruiting interventions include on-campus visits by prospects and off-campus recruiting visits by admissions representatives to high schools, community colleges, college fairs, etc. Remote recruiting interventions include postcards, brochures, email, text messages, and targeted social media.
<!-- QUESTIONS ABOUT THE RNL 2020 COST OF RECRUITING AN UNDERGRADUATE STUDENT REPORT, FIGURE 9

What does "recruitment publications" include?
what does "prospective student communications" include?
what does "event" include? can I assume this is on-campus events only cuz "travel" is a separate category?
what does "web services and digital advertising" include? Does this include social media?
-->

Each of these marketing/recruiting interventions -- and dozens of other interventions described in this book -- represent organizational behaviors that could be the subject of empirical research. However, the vast majority of research on enrollment management behaviors focuses on the final stages of the enrollment funnel, specifically which applicants are admitted and the use of financial aid "leveraging" to convert admits to enrollees [E.G.] <!-- [@RN3536; @RN3544; @RN3523; @RN4131; @RN2241; @RN3564; @RN1948] -->.  By contrast, the enrollment management industry expends substantial resources on marketing/recruiting activities that target earlier stages of the enrollment funnel [@RN4664] [GIVE SOME CONCRETE BUDGET NUMBERS INDICATING BALL PARK SPENDING?]. When we consider descriptions of enrollment management practice from investigative journalism [e.g., ], from enrollment management consulting firms and from this volume [e.g., ], it becomes apparent that -- with the notable exception of a small sociological case-study literature [e.g., ] -- the research community has ignores a great number of enrollment management practices that plausibly affect access to higher education. As a consequence, enrollment management remains an opaque industry to policymakers and the public. We argue that developing thoughtful policies about enrollment management depends on researchers collectively devloping empirical literatures that document enrollment management practices and evaluate the effects of these practices on opportunities for students.

Access to concrete data about marketing and recruiting behaviors has been a persistent barrier to empirical research. To the extent that marketing/recruiting behaviors are inconsistent with public relations rhetoric (e.g., "our commitment to access"), colleges and universities may have incentive to withhold data from researchers. Therefore, conventional data collection methods such as surveys and even interviews should be appraised with healthy skepticism. We argue that pushing forward an empirical research agenda on marketing/recruiting behaviors depends on different modes of data collection -- including public records requests and "data science" methodologies, such as web-scraping and streaming social media data -- that record actual behaviors (e.g., contracts, transactions, records of expenditure) as opposed to statements by colleges/universities about their behavior.

We present this chapter as an example of empirical research about marketing/recruiting behavior. This chapter analyzes off-campus recruiting visits by a sample of public research universities and a sample of selective private colleges and universities in 2017. We collected visit data by "scraping" URLs (e.g., "Coming to a neighborhood near you") once per week from college/university admissions websites and -- for the sample of public research universities -- by issuring public records requests. Our analyses to date have focused on public research universities. As described in more detail below, 12 of the 15 public research universities in our sample made far more out-of-state visits than in-state visits, a finding antithetical to the stated mission of most public research universities. These out-of-state visits focused on wealthy, predominantly white schools and communities. Additionally, most public research universities made a disproportionate number of out-of-state visits to private schools, but visited few in-state private schools.

Building on these findings, this chapter analyzes off-campus visits to private schools. In particular, we compare visits by public research universities to visits by selective private colleges and universities. The motivation for this public-private comparision is based on the idea that (out-of-state) visits to private high schools are antithetical to the stated mission of public research universities. By contrast, recruiting visits to private high schools are consistent with the mission of private colleges and universities. From this perspective, out-of-state visits to private schools by public research universities is one example of "privatization." A weakness of the privatization literature is that few empirical studies compare the public and private institutions with respect to particular behaviors associated with privatization [@RN4543]. Therefore, by comparing the (mostly out-of-state) visits to private high schools by public research universities to those of private colleges and universities, we hope to develop new insights about the privatization of public higher education.

Analyses are motivated by two, related research questions. First, which types of private high schools receive visits from which types of colleges and universities? Second, to what do extent colleges and universities visit private high schools that share similar characteristics with the college (e.g., geographic region, religious affiliation, academic reputation, and racial composition)? We use social network analysis (SNA) to analyze off-campus recruiting visits. The motivation for using SNA is that SNA privileges relationships between actorsQualitative scholarship on private high schools finds that college access is substantially determined by the quality of the relationship between a particular college and a particular private high school [@RN4321; @RN4407]. An off-campus recruiting visit is an indicator of a relationship between a college and a school that can be analyzed using standard tools of SNA.

[ADD ONE-PARAGRAPH OUTLINE?]

@@

# Network Methods and Data

We utilize network methods rather than conventional descriptive statistics for two, related reasons. First, social network analysis privileges the relationship between actors. The discussion of Stevens and Khan suggests that for private school students, "it is not just the quality of the students that gets them into college but the quality of the relationship between elite high schools and colleges" [@RN4407] [pg. 175]. Second, a visit from a college/university admissions counselor to a high school can be conceived as a "network tie" connecting two actors in a social network.

We briefly define essential social network analysis concepts and their operationalization in our analyses.  A social network consists of a set of actors -- referred to as "vertices" -- and the connections -- referred to as "edges" between these actors.  Whereas "one mode" networks consist of vertices of the same "type" (e.g., in a publication network each vertex is an author), "two mode" networks consit of vertices associated with one type of actor/entity having connections to vertices of another type. For example, an an actor-movie network consists of actors -- mode 1 -- who appear in movies -- mode 2 and an actor shares an edge with a movie if the actor appears in the movie.  The social network analyzed in this chapter is a two-mode network, where vertices consists of colleges/universities (mode 1) and high schools (mode 2) and an "edge" is defined as a visit from a college/university to a private high school [DELETE THIS SENTENCE OR THE REDUNDANT ONE IN BELOW PARAGRAPH].

Our analyses draw from analyses of "corporate board-director" networks because these are the most commonly analyzed two-mode networks in the social networks literature [e.g.] [@RN4663]. Board-director networks consist of directors (mode 1) who serve on organizational boards (mode 2). Our school-college network is a two-mode network composed of two sets of vertices, private high schools (mode 1) and colleges/universities (mode 2). Off-campus recruiting visits are edges (e.g., high school $i$ shares an edge with college $j$ if high school $i$ receives at least one visit from college $j$) and visits can only occur between a school-college pair (not a school-school pair or a college-college pair). Our school-college network is a two-mode network composed of high schools (mode 1) who are visited by colleges and universities (mode 2). This (weighted) network is represented as a school-by-college matrix (e.g., a $500 \times 40$ matrix if our network contains 500 high schools and 40 colleges/universities) in which matrix cell $a_{i,j}$ identifies the number of visits that high school $i$ received from college/university $j$. [SHOW AN EXCERPT OF THE TWO MODE ADJACENY MATRIX, LIKE 4 BY 3 OR SOMETHING] [DEFINE WEIGHTED VS. UNWEIGHTED SOMEWHERE]

Two-mode networks are ofen analyzed as one-mode networks [@RN2457]. @RN4663 states that "overlapping groups such as boards of directors form a two-mode membership network in which one can conceive of directors as nodes [vertices] connected by a tie of common board membership, or boards as nodes connected by a tie of one or more shared directors" [pg. xx]. Thus, a two-mode board-director network can be transformed into two one-mode networks. A weighted one-mode director network can be created as follows: create a director-by-director matrix (e.g., a $100 \times 100$ matrix if there are 100 directors in the network); two directors share an edge if they serve together on a board; matrix cell $a_{i,j}$ identifies the number of boards that include both director $i$ and director $j$ (e.g., $a_{i,j}=0$ if directors $i$ and $j$ do not serve on any of the same boards and $a_{i,j}=3$ if directors $i$ and $j$ do not serve on three boards together. Similarly, a one-mode board network is created based on a board-by-board matrix in which two boards share an edge if a director sits on both boards and matrix cell $a_{i,j}$ identifies the number of directors that serve on both board $i$ and board $j$. Following this approach, we create a one-mode high school network in which two private high schools share an edge if both receive a visit by the same college/university and cell $s_{i,j}$ of the school-by-school matrix $S$ indicates the number of colleges/universities that visited both school $i$ and school $j$ [SHOW EXCERPT OF ONE-MODE SCHOOL BY SCHOOL MATRIX, LIKE 3 BY 3 OR SOMETHING]. Similarly, we create a one-mode college/university network (herein the college network) in which two colleges share an edge if both colleges visited the same high school and cell $c_{i,j}$ of the college-by-college matrix $C$ indicates the number of private high schools that received at least one visit from both college $i$ and college $j$ [SHOW EXCERPT OF ONE-MODE COLLEGE BY COLLEGE MATRIX, LIKE 3 BY 3 OR SOMETHING].

A limitation of our data collection is that we collected off-campus visit data only from a convenience sample of X colleges and universities, as described above. Thus, for each college and university in our sample, we know all the private high schools they visited (assuming no measurement error). However, for each private high school in our sample, we only know visits they received from the set of X colleges/universities in our sample. Given this limitation, it may be useful to think of our data collection as resulting in X "ego networks." An "ego" is a single "focal" node (e.g., The University of Notre Dame). An ego network consists "of a focal node ("ego") and the nodes to whom ego is directly connected to (these are called "alters") plus the ties, if any, among the alters" [from http://www.analytictech.com/networks/egonet.htm#:~:text=Ego%20networks%20consist%20of%20a,form%20The%20human%20social%20network ]. For example, the University of Notre Dame ego network consists of all private high schools that received at least one visit from Notre Dame. Additionally, for each high school visited by Notre Dame, the ego network may include all colleges/universities in our sample that also visited the high school.

Figure X shows the ego network of all high schools visited by Notre Dame.

When analyzing ego networks, researchers often investigate the extent of "homophily" versus "heterophily" [@RN1722]. Homophily is the idea that two vertices are more likely to be connected if they have similar characteristics (e.g., two people are more likely to be freinds on Facebook if they have similar political ideologies). Heterophily, the opposite of homophily, is when pairs of connected vertices have different characteristics from one another. Figure X shows the University of Notre Dame ego network, with separate panels for vertex characteristics (academic reputation, geography, religious affiliation, and racial composition) and with vertex shapes (?colors?) to identify values of vertex characteristics [INSERT FIGURE]. Visually, we can see that ....[DESCRIBE BIG TAKEAWAYS].

Figure X augments the Notre Dame ego network by including the private colleges/universities in our sample that visited private high schools visited by Notre Dame [INSERT FIGURE]. ADD BRIEF TEXT ABOUT FIGURE?

@@

This section presents network analyses of visits by colleges and universities to private high schools. Analyses are motivated by two, related research questions. First, which types of private high schools receive visits from which types of colleges and universities? Second, to what extent do off-campus recruiting visits by colleges and universities to private high schools exhibit homophily versus heterophily with respect to the vertex characteristics of geographic region, religious affiliation, academic reputation, and racial composition?

We present results in three steps: First, visits by private colleges and universities; second, visits by public research universities; and, third, comparisons of visits by public research universities to visits by private colleges and universities.

## Visits by Private Colleges and Universities

CRYSTAL ADDS TABLES/FIGURES

## Visits by Public Research Universities

CRYSTAL ADDS TABLES/FIGURES

## Comparison of Public Versus Private Colleges and Universities

CRYSTAL ADDS TABLES/FIGURES
